---
title: "RMS Titanic"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task
in analyzing a dataset is to understand trends as they vary across these
different variables. Unless we're careful and thorough, we can easily
miss these patterns. In this challenge you'll analyze a dataset with a
small number of categorical variables and try to find differences among
the groups.

*Reading*: (Optional) [Wikipedia
article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics
define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----|----|----|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and
supporting files (`report_files/` folder) when you are done! Then submit
a link to Canvas. **Your Challenge submission is not complete without
all files uploaded to GitHub.**

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about
67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:

-   Class (Character: "1st", "2nd", "3rd", "Crew"
-   Sex: (Character: "Male", "Female")
-   Age: (Character: "Child", "Adult")
-   Survived (Character: "yes" or "no)
-   n (Double: number of people under a given group)

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

-   Write your observations here
-   Are there any differences?
    -   2224 - 2201 = 23
-   If yes, what might account for those differences?
    -   This is a small difference.
    -   We can only estimate because there are discrepancies between the
        original passenger and crew lists. Some factors that lead to
        this are misspellings, omissions, aliases, and failure to count
        contracted employees (such as musicians) as crew or passengers.
        Most estimates are based on the total capacity of the ship being
        around 2200.

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
titanic_survivors <- 
  df_titanic %>% 
  filter(Survived == "Yes")
titanic_survivors
titanic_survivors %>% 
  ggplot(aes(x = Class, y=n, fill = Sex)) +
  geom_col()
## TASK: Visualize counts against `Class` and `Sex`
```

**Observations**:

-   Generally, more female passengers survived. Most of the crew was
    probably male, which meant that more male crew survived.

# Deeper Look

<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for
understanding differences between groups. This is because the
differences we see in counts could be due to either the relative size of
the group OR differences in outcomes for those groups. To make
comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age`
group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
prop_titanic_survivors <- 
  df_prop %>% 
  filter(Survived == "Yes")
prop_titanic_survivors %>% 
  ggplot(aes(x = Class, y=Prop, fill = Sex)) +
  geom_col()
```

**Observations**:

-   Write your observations here.
    -   Generally, a bigger percentage of female occupants survived
        (compared to the overall female passenger population at that
        class) than male occupants of the same class.
-   Is there anything *fishy* going on in your plot?
    -   Although this plot shows percentages, the percentage itself is a
        bit unclear. It can be a percentage of the entire ship (which it
        is not) or a percentage of the same "type" of occupant. Although
        this shows the "likelihood" of a occupants surviving given their
        Class, Sex, and Age, it does not properly convey the proportion
        to the entire population.
    -   This is evident by the high number of female crew survivors,
        despite more male crew survivors in the raw data.
    -   Also, these proportions have not been adjusted to *percentages*
        that means that we are getting $>100\%$ percentages because the
        age groups are added together. If we summed `Prop`, we would get
        a $\sim 900\%$ survival rate. Thus, this chart is not useful and
        is prone to misunderstandings.

### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider
additional variables!

```{r q5-task}
total <- df_titanic %>% select(n) %>% sum()
df_titanic_total <- 
  df_titanic %>% 
  mutate(
    Prop = n / total
  )
prop_titanic_survivors_total <- 
  df_titanic_total %>% 
  filter(Survived == "Yes")
prop_titanic_survivors_total %>% 
  ggplot(aes(x = Sex, y=Prop, fill = Age)) +
  geom_col() +
  facet_grid(cols = vars(Class))
```

**Observations**:

-   This graph shows the true percentage of survivors compared to the
    total population of the Titanic occupants. We also separate the
    groups into adult and child for added clarity. If we added the
    percentages together, we would get $\sim32\%$, which is about the
    real survivor percentage.
-   If you saw something *fishy* in q4 above, use your new plot to
    explain the fishy-ness.
    -   In this graph, we were able to reach more believable proportion
        values. We also fixed the male-female crew survivor proportion.

# Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional
Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing
proportions is akin to non-dimensionalizing a quantity.
