---
title: "Aluminum Data"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-data03-pivot-basics
---

*Purpose*: When designing structures such as bridges, boats, and planes,
the design team needs data about *material properties*. Often when we
engineers first learn about material properties through coursework, we
talk about abstract ideas and look up values in tables without ever
looking at the data that gave rise to published properties. In this
challenge you'll study an aluminum alloy dataset: Studying these data
will give you a better sense of the challenges underlying published
material values.

In this challenge, you will load a real dataset, wrangle it into tidy
form, and perform EDA to learn more about the data.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics
define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and
supporting files (`report_files/` folder) when you are done! Then submit
a link to Canvas. **Your Challenge submission is not complete without
all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
```

*Background*: In 1946, scientists at the Bureau of Standards tested a
number of Aluminum plates to determine their
[elasticity](https://en.wikipedia.org/wiki/Elastic_modulus) and
[Poisson's ratio](https://en.wikipedia.org/wiki/Poisson%27s_ratio).
These are key quantities used in the design of structural members, such
as aircraft skin under [buckling
loads](https://en.wikipedia.org/wiki/Buckling). These scientists tested
plats of various thicknesses, and at different angles with respect to
the [rolling](https://en.wikipedia.org/wiki/Rolling_(metalworking))
direction.

# Loading and Wrangle

<!-- -------------------------------------------------- -->

The `readr` package in the Tidyverse contains functions to load data
form many sources. The `read_csv()` function will help us load the data
for this challenge.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/stang.csv"

## Load the data
df_stang <- read_csv(filename)
df_stang
```

Note that these data are not tidy! The data in this form are convenient
for reporting in a table, but are not ideal for analysis.

### **q1** Tidy `df_stang` to produce `df_stang_long`. You should have column names `thick, alloy, angle, E, nu`. Make sure the `angle` variable is of correct type. Filter out any invalid values.

*Hint*: You can reshape in one `pivot` using the `".value"` special
value for `names_to`.

```{r q1-task}
## TASK: Tidy `df_stang`
df_stang_long <-
  df_stang %>% 
  pivot_longer(
    names_to = c(".value", "angle"),
    names_sep = "_",
    values_to = "val",
    starts_with("E") | starts_with("nu"), 
    values_drop_na = TRUE
  ) %>% 
  filter(nu>0) %>% 
  mutate(angle = as.integer(angle))

df_stang_long
```

Use the following tests to check your work.

```{r q1-tests}
## NOTE: No need to change this
## Names
assertthat::assert_that(
              setequal(
                df_stang_long %>% names,
                c("thick", "alloy", "angle", "E", "nu")
              )
            )
## Dimensions
assertthat::assert_that(all(dim(df_stang_long) == c(26, 5)))
## Type
assertthat::assert_that(
              (df_stang_long %>% pull(angle) %>% typeof()) == "integer"
            )

print("Very good!")
```

# EDA

<!-- -------------------------------------------------- -->

## Initial checks

<!-- ------------------------- -->

### **q2** Perform a basic EDA on the aluminum data *without visualization*. Use your analysis to answer the questions under *observations* below. In addition, add your own *specific* question that you'd like to answer about the data---you'll answer it below in q3.

```{r q2-task}
summary(df_stang_long)
summarize(df_stang_long, mean(nu))
summarize(df_stang_long, n_distinct(alloy))
distinct(df_stang_long, angle)
distinct(df_stang_long, thick)
```

**Observations**:

-   Is there "one true value" for the material properties of Aluminum?
    -   There are many ways to characterize material properties. In this
        dataset, we use `nu`, which represents Poisson's ratio. Here, I
        computed the mean of the measured Poisson's Ratio.
-   How many aluminum alloys are in this dataset? How do you know?
    -   Only one type of aluminum alloy was tested in this dataset, as
        there was only one distinct alloy.
-   What angles were tested?
    -   Three angles were tested: 0, 45, 90
-   What thicknesses were tested?
    -   Four thicknesses were tested: .022, .032, .064, .081
-   Would it be fruitful to group by Young's modulus and see if there
    are any changes to Poisson's ratio?

## Visualize

<!-- ------------------------- -->

### **q3** Create a visualization to investigate your question from q2 above. Can you find an answer to your question using the dataset? Would you need additional information to answer your question?

```{r q3-task}
df_q3 <- 
  df_stang_long %>% 
  group_by(E) %>% 
  summarize(nu = mean(nu))
df_q3 %>% 
  ggplot(aes(x = E, y = nu)) + 
  geom_point()
```

**Observations**:

-   Although there is a rough correlation of higher Young's modulus
    resulting in a higher nu, the percent change is rather small, at
    \~20% change.
-   Additinally, this observation disregards any changes that may be due
    to another factor, such as thickness.

### **q4** Consider the following statement:

> "A material's property (or material property) is an intensive property
> of some material, i.e. a physical property that does not depend on the
> amount of the material."[2]

Note that the "amount of material" would vary with the thickness of a
tested plate. Does the following graph support or contradict the claim
that "elasticity `E` is an intensive material property." Why or why not?
Is this evidence *conclusive* one way or another? Why or why not?

```{r q4-vis}
## NOTE: No need to change; run this chunk
df_stang_long %>%

  ggplot(aes(nu, E, color = as_factor(thick))) +
  geom_point(size = 3) +
  theme_minimal()
```

**Observations**:

-   Does this graph support or contradict the claim above?
    -   This graph contradicts this claim because the "cluster" of the
        thickest samples are at the lowest end of both Young's modulus
        and Poisson's ratio. Thus, there must be some correlation with
        amount of material and these two properties, which in turn means
        that it cannot be an intensive property.
-   Is this evidence *conclusive* one way or another?
    -   Although there is a major grouping at the bottom, the other
        groupings are less conclusive. It looks like the Poisson's ratio
        of most samples with `thick<.081` are quite scattered across the
        `nu` axis. The Young's modulus looks like the highest points
        tend to be the least thick. However, the samples with
        `thick=.032` seem to have a lower mean Young's modulus then the
        sample with `thick=.064`. Additionally, we are able to see many
        outliers, despite the relatively small sample size. Thus, this
        evidence is inconclusive.

# References

<!-- -------------------------------------------------- -->

[1] Stang, Greenspan, and Newman, "Poisson's ratio of some structural
alloys for large strains" (1946) Journal of Research of the National
Bureau of Standards, (pdf
link)[<https://nvlpubs.nist.gov/nistpubs/jres/37/jresv37n4p211_A1b.pdf>]

[2] Wikipedia, *List of material properties*, accessed 2020-06-26,
(link)[<https://en.wikipedia.org/wiki/List_of_materials_properties>]
