---
title: "Challenge 13: Island Smoking"
author: "Christopher Nie"
date: 2025-05-11
output:
  github_document:
    toc: true
---

# Setup

```{r}
library(tidyverse)
```

# Data collection

### The Islands

I decided to study more Islands behavior. Specifically, I was interested in how the smoking behavior works. This observational study explores the various factors that may cause an individual islander to smoke more than others.

There may be some uncertainty because of the process of collecting data from the Islands by hand. The Islands has a total virtual population of \~40,000. I have sampled 270 points, or \~0.675% of the total population.

### Population and Sampling Plan

I wanted to study trends across all continents. I decided to use stratified sampling, where I randomly sampled 90 residents for each continent, keeping people sampled from the towns proportional to its size.

**Ironbard Population**

The sampled population of Ironbard is about 1% of the total population

| Town        | Total Population | Sampled Population |
|-------------|------------------|--------------------|
| *Hofn*      | 986              | 10                 |
| *Vardo*     | 3077             | 31                 |
| *Helvig*    | 1728             | 17                 |
| *Bjurholm*  | 1324             | 13                 |
| *Blonduos*  | 1067             | 11                 |
| *Helluland* | 868              | 9                  |
| **Total**   | 9050             | 91                 |

**Providence Population**

The sampled population of Providence is about 0.6% of the total population

| Town       | Total Population | Sampled Population |
|------------|------------------|--------------------|
| *Hayarano* | 1118             | 7                  |
| *Akkeshi*  | 1068             | 6                  |
| *Reading*  | 2024             | 12                 |
| *Nelson*   | 903              | 5                  |
| *Shinobi*  | 4822             | 29                 |
| *Biruwa*   | 1495             | 9                  |
| *Takazaki* | 1003             | 6                  |
| *Kiyobico* | 1301             | 8                  |
| *Arcadia*  | 966              | 6                  |
| **Total**  | 14700            | 88                 |

**Bonne Sante**

The sampled population of Bonne Sante is about 0.5% of the total population.

| Town      | Total Population | Sampled Population |
|-----------|------------------|--------------------|
| *Nidoma*  | 1055             | 5                  |
| *Talu*    | 821              | 4                  |
| *Pauma*   | 1038             | 5                  |
| *Valais*  | 905              | 5                  |
| *Kinsale* | 729              | 4                  |
| *Vaiku*   | 888              | 4                  |
| *Mahuti*  | 2649             | 13                 |
| *Eden*    | 1250             | 6                  |
| *Colmar*  | 5518             | 27                 |
| *Gordes*  | 1208             | 6                  |
| *Maeva*   | 862              | 4                  |
| *Riroua*  | 1210             | 6                  |
| **Total** | 18133            | 89                 |

### Quantity of Interest

An islander's life history displays "light smoker", "moderate smoker", "heavy smoker", and "nonsmoker." I use "nonsmoker" as the metric for judging how many times an islander started smoking – they can only quit if they started. For my quantity of interest, I was interested in exploring how age affects how many times someone starts smoking (and quits smoking).

### Covariates

Some covariates included are

-   Continent

-   Job

-   Net Worth

-   Sex

-   Smoker Depth (light, moderate, heavy)

### Question / Hypothesis

This study aims to explore the factors that cause an islander to smoke. My hypothesis is that age is the biggest determining factor, since the older the islander, the more chances they have to go to smoking.

# Analysis

### Setup

```{r}
filename <- "./data/smoking_islands_data_ironbard_providence.csv"
df_data_ironbard_providence <- read.csv(filename) 

df_data_ip <- df_data_ironbard_providence %>% 
  select(-X,-X.1,-X.2,-Notes)

filename <- "./data/smoking_islands_data_bonneSante.csv"
df_data_bonneSante <- read.csv(filename)

df_data_bs <- df_data_bonneSante %>% 
  select(-X,-X.1,-X.2,-X.3, -X.4)

df_data <- full_join(
  df_data_ip, 
  df_data_bs, 
  by = c("continent","age", "job", "Net.wrth", "Gender", "first_smoke_age", "nonsmoke_instance", "smoker_depth", "town")) %>% 
  mutate(first_smoke_age = na_if(first_smoke_age, 0))
tail(df_data) 
```

Dataframe columns are:

-   Continent

-   Town

-   Age

-   job

-   Net worth

-   Gender

-   first_smoke_instance: first case of "light smoker"

-   nonsmoke_instance: number of instances of "nonsmoker"

-   smoker_depth: shows the heaviest level of smoking recorded in their life history

-   smoke_instance: shows the number of instances of "\_\_\_ smoker"

-   light_Instance: shows the number of instances of "light smoker"

-   moderate_instance: shows the number of instances of "moderate smoker"

-   heavy_instance: shows the number of instances of "heavy smoker"

### Smoking and Age

#### Smoking Cycles

```{r}
df_data %>% 
  filter(!is.na(first_smoke_age)) %>% 
  ggplot(aes(x = first_smoke_age, color = continent)) +
  geom_density() +  # Overlapping densities
  labs(
    title = "Distribution of First Smoking Age by Continent",
    x = "Age at First Smoking",
    y = "Density"
  ) +
  theme_minimal() +
  scale_color_brewer(palette = "Dark2")
```

We see that the most common starting age is between 18 to 21, with 19 being the mode for all continents. Interestingly, Ironbard has a second mode at 21. This starting age may suggest that someone younger (18-24) may have only gone through one or two smoking cycles. To do this, I filtered the data by age and checked the reoccurrence of smoking.

```{r}
df_data %>% 
  filter(age<25) %>% 
  ggplot(aes(x = nonsmoke_instance, color = continent))+
  geom_density() + 
  theme_minimal() +
  labs(
    title = "Count of Smoking Instances (younger than 25)",
    x = "Smoking instances",
    y = "Density"
  )
```

We can see that, in general for most continents, the densities are right skewed, showing that most people in this age group have fewer smoking cycles. We can also see that Ironbard seems to have a llot of nonsmokers, at almost 50%. Let's see if this carries over to older islanders (\>35).

```{r}
df_data %>% 
  filter(age>30) %>% 
  ggplot(aes(x = nonsmoke_instance, color = continent))+
  geom_density() +
  theme_minimal() +
  labs(
    title = "Count of Smoking Instances (older than 35)",
    x = "Smoking Instances",
    y = "Density"
  )
```

We can see that the median and mode shift to the right. The mode of Providence is more concentrated around 8. Bonne Sante is slightly less concentrated around 6, and Ironbard is spread out across 5 to 11. We can also see this trend if we plotted everything on a scatterplot.

```{r}
df_data %>% 
  mutate(smoker_depth = case_when(
    smoker_depth == 0 ~ "nonsmoker",
    smoker_depth == 1 ~ "light",
    smoker_depth == 2 ~ "moderate",
    smoker_depth == 3 ~ "heavy",
    TRUE ~ as.character(smoker_depth)
  )) %>% 
  mutate(smoker_depth = forcats::fct_relevel(smoker_depth, "nonsmoker", "light", "moderate", "heavy")) %>% 
  ggplot(aes(x = age, y = nonsmoke_instance, color = smoker_depth)) + 
  geom_point()+
  theme_minimal() +
  labs(
    title = "Age vs Smoking cycles",
    x = "Age",
    y = "Smoking cycles", 
    color = "Smoker Level"
  )
```

This graph shows the general trend that smoking cycle increases with age. However, it also shows another metric used to define the smokers: Smoker Level. And unlike smoking cycles, it does not appear to be directly correlated with age. It does, however, appear somewhat correlated to smoking cycles.

#### Smoker Level

To explore smoker level, I tried plotting it directly against `Age` and `nonsmoke_instance`. I also looked at the effects of `first_smoke_age`. None of these graphs revealed any trends. I ended up filtering by age and looking at their distributions.

```{r}
df_data %>%
  mutate(smoker_depth = case_when(
    smoker_depth == 0 ~ "nonsmoker",
    smoker_depth == 1 ~ "light",
    smoker_depth == 2 ~ "moderate",
    smoker_depth == 3 ~ "heavy",
    TRUE ~ as.character(smoker_depth)
  )) %>% 
  mutate(smoker_depth = forcats::fct_relevel(smoker_depth, "nonsmoker", "light", "moderate", "heavy")) %>% 
  filter(nonsmoke_instance<=3) %>% 
  ggplot(aes(x = smoker_depth, fill = continent))+
  geom_bar(position = "dodge") + 
  theme_minimal() +
  labs(
    title = "Count of Smoker Levels (3 or less smoking cycles)",
    x = "Smoking Levels",
    color = "Continent"
  )

df_data %>%
  mutate(smoker_depth = case_when(
    smoker_depth == 0 ~ "nonsmoker",
    smoker_depth == 1 ~ "light",
    smoker_depth == 2 ~ "moderate",
    smoker_depth == 3 ~ "heavy",
    TRUE ~ as.character(smoker_depth)
  )) %>% 
  mutate(smoker_depth = forcats::fct_relevel(smoker_depth, "nonsmoker", "light", "moderate", "heavy")) %>% 
  filter(nonsmoke_instance>=7) %>% 
  ggplot(aes(x = smoker_depth, fill = continent))+
  geom_bar(position = "dodge") + 
  theme_minimal() +
  labs(
    title = "Count of Smoker Levels (7 or more smoking cycles)",
    x = "Smoking Levels",
    color = "Continent"
  )
```

From these graphs, we can see that the mode of the graph definitely shifts to the right, suggesting that smoking cycles might be a valid determiner of smoker level. Noticeably, there are still some heavy smokers who have only been through three smoking cycles. This means that by extension, smoker level is somewhat related to age, if only by proxy of smoking cycle. This appears to be valid for all continents.

### Other factors

Some other factors I considered were

-   Sex

-   Employment status

-   Net worth

#### Sex

This factor requires us to normalize our data, as we have 120 female datapoints and 134 male datapoints.

```{r}
df_data %>% 
  filter(Gender != "") %>% 
  mutate(is_smoker = (smoker_depth!=0)) %>% 
  group_by(Gender, is_smoker) %>% 
  summarise(count = n()) %>% 
  mutate(prop = count / sum(count)) %>% 
  ggplot(aes(x = is_smoker, y = prop, fill = Gender)) +
  geom_col(position = "dodge") + 
  theme_minimal() +
  labs(
    title = "Smoker vs non-smoker (Sex)",
    x = "Has smoked before",
    y = "Proportion",
    fill = "Sex"
  )
```

It appears that proportionally, there are more female nonsmokers than male nonsmokers, and more male smokers than female smokers. However, the difference in proportions is \~2%, which may just be an artifact of the sampling process and probably masked by uncertainty. In fact, the actual count was 9 female nonsmokers and 7 male nonsmokers. Given our scale and sampling limitations, we can only say that there is no meaningful difference between genders regarding smoking.

However, this just shows whether or not an islander is likely to smoke at all. On the islands, it appears that smoking is such widespread behavior that it is rather uncommon to not have smoked at all. Thus, we can look at whether sex is correlated with *smoking seriousness –* how likely they are to go beyond a light smoker.

```{r}
df_data %>% 
  filter(Gender != "") %>% 
  mutate(is_smoker = (smoker_depth>1)) %>% 
  group_by(Gender, is_smoker) %>% 
  summarise(count = n()) %>% 
  mutate(prop = count / sum(count)) %>% 
  ggplot(aes(x = is_smoker, y = prop, fill = Gender)) +
  geom_col(position = "dodge") + 
  theme_minimal() +
  labs(
    title = "non/light smoker vs moderate/heavy smoker (Sex)",
    x = "Is a Moderate or Heavy Smoker",
    y = "Proportion",
    fill = "Sex"
  )
```

Here, we can see a statistically significant difference. There are more female light smoker-only islanders than male light smoker-only islanders. However, for both groups, \>50% end up as moderate or heavy smokers, although the proportion for females are much closer to 50-50. We can also refer to our original metric, number of smoke cycles.

```{r}
df_data %>% 
  filter(Gender != "") %>% 
  mutate(is_smoker = (nonsmoke_instance>5)) %>% 
  group_by(Gender, is_smoker) %>% 
  summarise(count = n()) %>% 
  mutate(prop = count / sum(count)) %>% 
  ggplot(aes(x = is_smoker, y = prop, fill = Gender)) +
  geom_col(position = "dodge") + 
  theme_minimal() +
  labs(
    title = "More than five smoking cycles (Sex)",
    x = "Has gone through more than five smoking cycles",
    y = "Proportion",
    fill = "Sex"
  )
```

This graph shows that most female islanders undergo less than five smoking cycles, and most male islanders undergo more than five smoking cycles.

To sum up this section, sex does not meaningful impact whether an islander has smoked before. However, it does affect how deep the smoking goes – whether it is smoking level or number of smoking cycles.

#### Employment status

This factor requires us to normalize the data, so that we can properly see whether being employed significantly impacts the chance to smoke.

```{r}
df_data %>% 
  filter(Gender != "") %>% 
  mutate(
    is_smoker = (smoker_depth!=0),
    is_employed = (job!="")
    ) %>% 
  group_by(is_employed, is_smoker) %>% 
  summarise(count = n()) %>% 
  mutate(prop = count / sum(count)) %>% 
  ggplot(aes(x = is_smoker, y = prop, fill = is_employed)) +
  geom_col(position = "dodge") + 
  theme_minimal() +
  labs(
    title = "Smoker vs non-smoker (Employment)",
    x = "Has smoked before",
    y = "Proportion",
    fill = "Employed"
  )
```

We see a pretty similar situation to the *smoker vs non-smoker (sex)* graph. In this case, the proportion difference is \~1%, which is even smaller. Thus, there is no statistically significant difference in tendency to smoke whether an islander is employed or not employed.

However, just like the sex section, we can look at whether employment affects the level of smoking.

```{r}
df_data %>% 
  filter(Gender != "") %>% 
  mutate(
    is_smoker = (smoker_depth>1),
    is_employed = (job!="")
    ) %>% 
  group_by(is_employed, is_smoker) %>% 
  summarise(count = n()) %>% 
  mutate(prop = count / sum(count)) %>% 
  ggplot(aes(x = is_smoker, y = prop, fill = is_employed)) +
  geom_col(position = "dodge") + 
  theme_minimal() +
  labs(
    title = "non/light smoker vs moderate/heavy smoker (Employment)",
    x = "Is a Moderate or Heavy Smoker",
    y = "Proportion",
    fill = "Employed"
  )
df_data %>% 
  filter(Gender != "") %>% 
  mutate(
    is_smoker = (nonsmoke_instance>5),
    is_employed = (job!="")
    ) %>% 
  group_by(is_employed, is_smoker) %>% 
  summarise(count = n()) %>% 
  mutate(prop = count / sum(count)) %>% 
  ggplot(aes(x = is_smoker, y = prop, fill = is_employed)) +
  geom_col(position = "dodge") + 
  theme_minimal() +
  labs(
    title = "More than five smoking cycles (Employment)",
    x = "Has gone through more than five smoking cycles",
    y = "Proportion",
    fill = "Employed"
  )
```

These graphs show that, similar to sex, employment status does not meaningful impact whether an islander has smoked before. However, it does appear that unemployed islanders tend to become moderate/heavy smokers, and also tend to go through more smoking cycles. In fact, most employed islanders do not go through more than 5 smoking cycles. However, this connection is a bit trickier, as employment situation can change(whereas Islander sex presumably cannot). Thus, it is possible that employment status is merely a symptom of another (perhaps behavioral) characteristic that determines whether an islander is prone to smoking. Regardless, the sheer statistical difference showcased in this final graph reveals that there might be something worth exploring.

Another possible reason for this difference might be simply be age. Most older islanders are unemployed, and as established previously, older islanders tend to undergo more smoking cycles.

#### Net worth

```{r}
df_data %>% 
  mutate(smoker_depth = case_when(
    smoker_depth == 0 ~ "nonsmoker",
    smoker_depth == 1 ~ "light",
    smoker_depth == 2 ~ "moderate",
    smoker_depth == 3 ~ "heavy",
    TRUE ~ as.character(smoker_depth)
  )) %>% 
  mutate(smoker_depth = forcats::fct_relevel(smoker_depth, "nonsmoker", "light", "moderate", "heavy")) %>% 
  ggplot(aes(x = Net.wrth, y = nonsmoke_instance, color = smoker_depth)) + 
  geom_point()+
  scale_x_log10() + 
  theme_minimal() +
  labs(
    title = "Net Worth vs Smoking cycles",
    x = "Net Worth",
    y = "Smoking cycles", 
    color = "Smoker Level"
  )
```

This graph shows that Net Worth and Smoking cycles may be negatively correlated. As an islander gets richer, their smoking cycles goes down. Notably, there are nonsmokers all along the net worth line. However, when I was making the dataset, I realized that the net worth of older islanders tended to be lower. Thus, it is possible that this correlation might go back to age. Thus, I made this plot again, but changed the color scale to be age.

```{r}
df_data %>% 
  ggplot(aes(x = Net.wrth, y = nonsmoke_instance, color = age)) + 
  geom_point()+
  scale_x_log10() + 
  theme_minimal() +
  labs(
    title = "Net Worth vs Smoking cycles",
    x = "Net Worth",
    y = "Smoking cycles", 
    color = "Smoker Level"
  )
```

We can see that the the higher points tended to just be those who were older. As we go downwards on the smoking cycles axis, we see that the color becomes darker, signifying the data points being younger. The nonsmokers at the bottom are simply the 20 years old who haven't started smoking yet.

Thus, coming back to my hypothesis: *All smoking factors go to age.*

### Heavy Smokers

I was interested in the mechanism behind the heavy smokers. When I was collecting the data for Bonne Sante, I realized that I could simply record the instances of `light smoker`, `moderate smoker` and `heavy smoker`. This would let me see whether the number of `light smoker` instances was correlated with whether an islander would become a `moderate smoker`, and whether the number of `moderate smoker` instances was correlated with an islander's tendency to become a `heavy smoker`.

```{r}
df_data %>% 
  filter(!is.na(moderate_instance)) %>% 
  mutate(smoker_depth = case_when(
    smoker_depth == 0 ~ "nonsmoker",
    smoker_depth == 1 ~ "light",
    smoker_depth == 2 ~ "moderate",
    smoker_depth == 3 ~ "heavy",
    TRUE ~ as.character(smoker_depth)
  )) %>% 
  mutate(smoker_depth = forcats::fct_relevel(smoker_depth, "nonsmoker", "light", "moderate", "heavy")) %>% 
  ggplot(aes(x = light_instance, y = moderate_instance, color = smoker_depth))+
  geom_point() + 
  theme_minimal() +
  labs(
    title = "Light smoker instance vs Moderate smoker instance",
    x = "Light smoker instance",
    y = "Moderate smoker instance", 
    color = "Smoker Level"
  )

df_data %>% 
  filter(!is.na(moderate_instance)) %>% 
  mutate(smoker_depth = case_when(
    smoker_depth == 0 ~ "nonsmoker",
    smoker_depth == 1 ~ "light",
    smoker_depth == 2 ~ "moderate",
    smoker_depth == 3 ~ "heavy",
    TRUE ~ as.character(smoker_depth)
  )) %>% 
  mutate(smoker_depth = forcats::fct_relevel(smoker_depth, "nonsmoker", "light", "moderate", "heavy")) %>% 
  ggplot(aes(x = moderate_instance, y = heavy_instance, color = smoker_depth))+
  geom_point()+
  geom_point() + 
  theme_minimal() +
  labs(
    title = "Moderate smoker instance vs Heavy smoker instance",
    x = "Moderate smoker instance",
    y = "Heavy smoker instance", 
    color = "Smoker Level"
  )
```

In general, we can see that the instances increase with each other. This is because in order for an islander to be a moderate smoker, they must first be a light smoker, and in order for an islander to be a heavy smoker, they must first be a moderate smoker. Additionally, in order to become a nonsmoker, they must drop down to light smoker before they can become a nonsmoker. This means that in general, for any given islander, there must be more light smoker instances than moderate smoker instances, and more moderate smoker instances than heavy smoker instances. It appears that most islanders follow this circle, where they will drop all the way down before starting another smoking cycle. However, there are exceptions. For example, there is a point (2,5) on the *Light smoker instance vs Moderate smoker instance* graph. This is because the the person became a light smoker, then a moderate smoker, then bounced between moderate and heavy smoker. However, this seems to be the only data point with this characteristic.

Another way we can look at the graph is playing the game of probabilities. Just like how age determines how many smoking cycles an islander goes through, more light smoker instances increases the probability that eventually one of them becomes a moderate smoker instance.

# Conclusion

From the analysis, we can see that, of the visible factors collected, age is most heavily correlated to the number of smoking cycles. Smoking cycles seem to be correlated with smoker level. Some of the other factors we looked at, such as employment or economic status, could have simply been proxies for age. We were also able to see how the light smoker, moderate smoker, and heavy smoker state were connected. However, the actual stochastic model behind it is still a mystery. Perhaps time in each state is important. Unfortunately, that would have been too difficult to collect and analyze, especially as many islanders went through multiple smoking cycles.

My sample of 90 from each continent resulted in 270 islanders. The total population of the islands is 40000. This gives me a 6% confidence interval at a 95% confidence level. [For most research conducted at the population level, researchers look for at least 385 samples.](https://www.geopoll.com/blog/sample-size-research/#:~:text=As%20demonstrated%20through%20the%20calculation,are%20often%20used%20in%20research.) Thus, raising the sample to this level would improve our confidence interval.

Some questions remain. Family history, such as genetics or parent behavior, may affect the smoking behavior of any children they have. In other words, is smoking "infectious" – can an islander make the people around them more likely to smoke if they smoke? Another question that remains is how a light smoker becomes a moderate smoker, and a moderate smoker a heavy smoker. Is it the "same" randomness across all islanders, or are some islanders inherently more susceptible to it? The stance I have taken is that it is the same across all islanders, but the older islanders have more time for the probability hammer to fall. However, if this were completely true, then why are there islanders who have 4 heavy smoker instances and 4 moderate smoker instances when there are islanders who have 9 moderate smoker instances but only 1 heavy smoker instances? Why are there islanders that have hit 19 light smoker instances but only 1 moderate smoker instance? Or 11 light smoker instances but no moderate smoker instances? One factor I considered collecting was time in each state, but it was unfeasible for me to collect that data. It would definitely be an interesting direction to explore.

Finally, the biggest question of all remains: why do more than 90% of islanders smoke at least once in their life?
